From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study

### 2025-08-12 draft_v1/draft_v2.qmd
draft update
vitor_feedback_2nd

```{r, include=FALSE, cache=TRUE}
# Load the dataset 
dat <- read_sav("../../data/HELIUS_LEONIE.sav")

@@

### Depression as sum score
The sum score graphs provide a high-level summary of how precariousness factors collectively influence overall depression severity, focusing on aggregated relationships.
@fig-sum illustrates the causal relationships between precariousness factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*), as identified by the FCI algorithm.
As described in [@sec-analysis], dashed edges indicate ties, and the edge endpoint at which the tie occurs is marked by a bold horizontal bar.

The causal graph in panel (a) highlights key pathways in the relationships between precariousness factors and depression (*PHQsum*). Employment precariousness (*P.emp*) and social precariousness (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role. While *P.emp* and *P.soc* are generally not recognized as causes of other precariousness factors, *S.fin* emerges as a potential cause, as indicated by its circle edge endpoint. This pattern is further supported by the edge-type proportion matrix shown in panel (b), which aggregates edge inference outcomes across bootstrapped datasets.

The matrix plot can be interpreted such that the symbol in matrix[$i$, $j$] represents the relationship $i -~[symbol] \  j$. For example, if matrix[$i$, $j$] is > and matrix[j, i] is $\circ$, then the inferred relationship is $i \circarrow j$.  In the table, different edge types are represented by distinct colors: light gray for the absence of an edge, blue for circles, green for arrowheads, and coral for arrowtails. The colors are blended based on the proportion of each edge type, with higher proportions increasing the opacity of the corresponding color, making dominant symbols more visually prominent.

@@

The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI algorithm, with one exception: CCI does not identify *S.fin* as a cause of depression. Additionally, CCI introduces greater uncertainty in edge directions, with more ties mainly between circle and arrowhead endpoints, and tends to favor arrowheads more frequently than FCI. Despite these differences, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm, supporting the key pathways.

The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI analysis, with one key distinction: CCI shows greater ambiguity in the directional relationship between *S.fin* and *PHQsum*: CCI assigns an arrowhead from *S.fin* to *PHQsum* in roughly 50% of bootstraps, while the remaining cases display an undirected (circle) endpoint.
Overall, CCI introduces greater uncertainty in edge orientations relative to FCI, yielding more undirected endpoints and more frequent bidirectional (confounded) edges, which in turn contribute to a slightly higher overall count of arrowheads when directions are inferred. 
Despite this, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm. For example, both algorithms identify directional pathways from the stressors (*S.rel*, *S.fin*) to *P.soc*, and suggest that *P.soc* is not a cause of *PHQsum*, but more likely affected by it. *P.emp* is not inferred to cause any of its neighboring variables (*P.hou*, *PHQsum*, *S.fin*), and instead appears to be affected by them. *P.hou*, in turn, is connected to *P.emp* and *S.fin* with circle endpoints, indicating that it may be a potential source of influence rather than an outcome. These consistent pathways underscore the robustness of the core relational architecture across algorithms amid variation in directional confidence.

@@

@@

Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms. Unlike the sum score graph, which aggregates all symptoms into a single measure, the symptom-level graph reveals the heterogeneity in how specific symptoms (*con* (concentration), *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood)) are influenced by, and in turn influence, different forms of precariousness.

The symptom-level graph in @fig-sym-1 reveals a more complex and interconnected structure than the sum score graph, reflecting the strong interdependence among symptoms and suggesting much presence of latent confounders, as indicated by numerous bidirectional edges. As before, dashed edges denote ties, with specific ties marked by bold horizontal bars. 
In @fig-sym-2, the proportion matrix provides more insight into these directionalities. Here, opaque green dominates most symptom-to-symptom connections, indicating that arrowheads are the most frequently inferred edge type. However, blue regions, which correspond to circle endpoints, are particularly common for *anh*, *slp*, *ene*, and *sui*, reflecting uncertainty in the causal direction among these symptoms. Notably, the only arrowtail connection appears between *dep* and *anh*, suggesting that anhedonia is a potential cause of depressed mood.

@@

This discrepancy likely arises from a combination of factors. 
As the analysis moves from aggregated to disaggregated variables, the number of CI tests increases substantially, leading to a loss of statistical power. With more variables being conditioned on, the effective sample size per test diminishes, making it harder to detect subtle dependencies. Over-conditioning—controlling for too many variables—can also obscure real relationships, resulting in false negatives.
Beyond these statistical power issues, the underlying assumptions of the two CI tests contribute to their divergence. RCoT is a nonparametric test that captures both linear and nonlinear dependencies without assuming a specific distributional form. However, it is more conservative and requires larger sample sizes, especially in high-dimensional settings. In contrast, the Gaussian CI test is based on partial correlations and assumes linear Gaussian relationships, making it more permissive and more likely to detect weak or spurious associations.
As a result, it is common to see edges where Gaussian CI suggests a directional relationship, while RCoT indicates no edge. In the proportion matrix, this appears as a 50–50 tie between absence and arrowhead symbols. 
<!-- This systematic divergence illustrates the broader challenge of causal discovery in high-dimensional data: different statistical assumptions can lead to different conclusions about whether—and in which direction—variables are causally related. -->

@@

Other precariousness factors exhibit more uncertainty but still notable relationships. *S.rel* shows weak but existing connections with *slp*, *glt*, and *app*, though directionality remains unclear in many cases. *P.emp* also connects to symptoms like *slp*, *mot*, and *con*, but, like *S.rel*, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
In line with the aggregated graph in @fig-sum, *P.hou* does not appear to have any direct relationship with depression symptoms but consistently shows associations with *P.emp* and *S.fin*. The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with *S.fin* exhibiting the strongest tendency to influence other precariousness factors.

Among depressive symptoms, *dep*, *glt*, and *slp* appear to be the most connected to precariousness factors, while *P.soc* has the most connections with symptoms, predominantly as a recipient rather than a driver of influence. Within the symptom network, *slp*, *sui*, and *anh* emerge as causally influential symptoms, as they exhibit more outgoing arrows compared to other symptoms. On the other hand, *con*, *mot*, *dep*, and *glt*, despite having high connectivity, predominantly receive incoming arrows, indicating they are more likely effects rather than causes.
Considering both symptom-precariousness connections and symptom-level dynamics, *slp* emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network. This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precariousness. On the other hand, *P.soc* serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precariousness, reinforcing a self-sustaining dynamic between depression and precarious conditions.

@@

Unlike the disaggregated graph in @fig-sym, the aggregated symptom-precariousness graph (@fig-presum-1) shows more certainty in causal directions, with only a few edges resulting in ties, all of which involve the overall precariousness factor. This is even more evident in the proportion matrix (@fig-presum-2), where symptom-to-symptom interactions almost fully converge to a single edge type. Additionally, all symptom interactions are connected either through bidirectional edges or a combination of an arrowhead and a circle, suggesting a strong presence of latent confounders.

Examining the symptom-precariousness connections, we observe a stronger overall trend of depressive symptoms influencing precariousness, rather than the other way around. Specifically, *anh*, *dep*, and *slp* show a high probability of causing precariousness, whereas *app*, *glt*, and *mot* exhibit more uncertainty in directionality, often resulting in circle endpoints. This pattern suggests that when precariousness factors are aggregated, the dominant causal flow is from depressive symptoms to precariousness, rather than precariousness driving depression. Particularly, *dep* emerges as the strongest predictor of precariousness, exhibiting the highest proportion of arrowtails, consistent with findings from the disaggregated analysis.

@@

Overall, the aggregated precariousness graph supports the key findings from both the disaggregated analysis and the depression sum score analysis, providing greater certainty in several important patterns. The results highlight the significant role of latent confounders in symptom-to-symptom interactions. While certain stressor-related precariousness factors, particularly *S.fin* and *S.rel*, appear to causally influence depressive symptoms, other precariousness factors are more likely driven by depression rather than acting as primary causes, suggesting that precariousness may function more as a consequence of depression than its source. Some symptoms, particularly *glt*, *slp*, and *dep*, appear to influence precarious conditions while also occupying central roles within the symptom network. Among them, *slp* and *glt* emerge as influential symptoms within the depression network, whereas *con* and *dep*, despite their high connectivity, behave more as outcomes rather than primary drivers. Together, these findings suggest that depression—especially symptoms like sleep disturbances and guilt—may contribute to sustaining precarious conditions, possibly forming a reciprocal cycle between depression and precariousness.

<!-- As before, black solid lines indicate edges agreed upon by both FCI and CCI, while gray dashed lines represent edges where the two algorithms disagree. Circles at the endpoints of gray dashed edges denote unresolved directionality. Additionally, navy dashed edges highlight relationships that appear exclusively in graphs generated using Gaussian CI tests. -->
<!-- Compared to @fig-sym, this analysis reveals greater disagreement between FCI and CCI regarding edge directionality, as evidenced by more dashed edges. FCI tends to estimate more undirected edges (circles), while CCI more frequently assigns directed arrows between symptom-symptom relationships. For a detailed breakdown, see [@sec-propmatrix], which shows the proportion of symbols produced by each algorithm. -->
<!-- When examining overall precarity, we observe that more symptoms are now connected to the precarity score (*Precarity*) compared to the individual precarity factor analysis in @fig-sym. In addition to *glt*, *slp*, *anh*, and *sui*, which are previously shown to be connected to individual precarity factors, *mot*, *app*, and *dep* are now identified as having causal relationships with overall precarity. Furthermore, when using the Gaussian CI test, con is also shown to be causally related to overall precarity. Notably, all these connections are directed towards precarity (o→), indicating that overall precarity is not the cause of symptoms. The exception is the bidirectional edge between *sui* <-> *Precarity*, which suggests a connection via latent confounders. -->
<!-- FCI predominantly estimates directed arrows (→) for symptom-precarity relationships, suggesting that symptoms are driving precarity. With fewer variables (reduced from 14 to 9), this approach appears to have slightly more statistical power, enabling algorithms to estimate relationships with greater confidence compared to the previous analysis involving individual symptoms and individual precarity factors. -->

```{r}
#| layout-nrow: 2
#| label: fig-presum
#| fig-cap: "Resulting graph of precariousness sum score and individual depression symptoms using FCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 60%

@@

To assess how internal structure influences the system’s responsiveness to external support, we simulate a set of linear models that are consistent with the data but vary in the assumed strength of influence from precariousness to depression ($\alpha_{DP}$).
For each model, we generate a virtual population of 1,000 individuals and simulate an external intervention that reduces financial stress ($S$), which acts as a shared input across individuals. This intervention defines the control input, while different values of $\alpha_{DP}$ represent alternative internal feedback scenarios.
We examine system behavior in two complementary ways: (1) by analyzing average levels of depression and precariousness at selected time points following intervention onset, and (2) by simulating full temporal trajectories under sustained intervention followed by recovery, using two representative cases at the extremes of the $\alpha_{DP}$ range.

@fig-intervention summarizes how variation in $\alpha_{DP}$ shapes system-level responses. Panel A shows the mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, with the x-axis indicating the strength of external support, defined as the proportional reduction of $S$ toward its minimum observed value.
Each line represents a different scenario with a different $\alpha_{DP}$, color-coded from low (purple) to high (yellow).

In the early phase of intervention ($t = 100$), depression and precariousness exhibit contrasting patterns. Depression declines more rapidly in scenarios with lower $\alpha_{DP}$, where its dynamics are driven more directly by the external stressor $S$. 
As $\alpha_{DP}$ increases, $D$ becomes more dependent on precariousness ($P$), which itself adjusts more slowly—delaying the depressive response.
Precariousness, by contrast, shows slightly faster early reductions in scenarios with higher $\alpha_{DP}$. This reflects a shift in influence: as $\alpha_{DP}$ increases, $P$ becomes more directly responsive to $S$ (via increasing $\alpha_{PS}$) and less influenced by $D$, allowing it to adjust more promptly to changes in the stressor.

Despite these short-term differences, all scenarios ultimately converge to similar steady-state levels under sustained intervention. By time step 300, convergence is already substantial—though some variation remains, particularly in $P$, where gradients between models are still visible. By step 500, trajectories have nearly fully aligned across all configurations, consistent with the shared equilibrium structure of the linear system. This convergence highlights that the key distinctions lie not in whether intervention is effective, but in how quickly and through which pathways change unfolds.

Panel B shows full time trajectories for two representative cases: one with minimal internal feedback ($\alpha_{DP} = 0$) and one at the upper admissible limit ($\alpha_{DP} = 0.65$). In both, intervention begins at step 300 and ends at step 1500. When $\alpha_{DP} = 0$, depression responds rapidly to the reduction in stress and rebounds shortly after support is withdrawn. In contrast, when $\alpha_{DP}$ is high, $D$ responds more gradually and recovers more slowly—reflecting its increased dependence on the slower-changing $P$.
Precariousness declines in both scenarios during intervention and returns to baseline afterward, but with a slightly sharper drop in the higher $\alpha_{DP}$ condition. This occurs because $P$ becomes more directly responsive to $S$, as $\alpha_{PS}$ increases with $\alpha_{DP}$.

@@

```{r}
#| label: fig-intervention
#| fig-cap: >
#|   System response to external intervention across variations in internal structure.
#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across different values of $\alpha_{DP}$ and intervention strengths.
#|   Models with lower $\alpha_{DP}$ show faster early reductions in $D$, while higher $\alpha_{DP}$ leads to more immediate declines in $P$.
#|   **Panel B**: Full time trajectories for two models—one with no coupling ($\alpha_{DP} = 0$) and one with strong coupling ($\alpha_{DP} = 0.65$)—illustrating system response and recovery following intervention onset (step 300) and withdrawal (step 1500).
#|   Depression responds more gradually and persistently in the high $\alpha_{DP}$ model, whereas precariousness adjusts more quickly.*

@@

<!-- Summmary & Interpretation of Findings -->
This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precariousness interact—and how their internal structure modulates responsiveness to intervention. Using cross-sectional data from the HELIUS study, we applied cycle-capable discovery algorithms to estimate plausible directional relationships among multiple domains of precariousness factors and depressive symptoms. By analyzing both aggregate constructs and individual variables, we aimed to capture patterns that may be masked in either extreme of abstraction (i.e., purely aggregate or disaggregated analyses).

The causal analysis consistently identified financial stress as a key plausible driver of depression, appearing across both aggregated and disaggregated models. 
At the symptom level, several nodes—sleep disturbance, depressed mood, guilt, and anhedonia—emerged as especially sensitive to external influences. Among these, depressed mood played a particularly central role: it was influenced by other symptoms and, in turn, showed directional influence on social precariousness, suggesting a feedback mechanism wherein worsening mood fosters social withdrawal, which may subsequently exacerbate depressive symptoms. 
The CCI algorithm further suggested a potential cyclic link between depressed mood and overall precariousness, hinting that symptom and precariousness factors may co-evolve in mutually reinforcing ways.
Symptom sleep disturbance also played a key initiating role. It showed directed connections to both social and employment precariousness and exhibited many outgoing edges within the symptom network. This suggests the symptom sleep disturbance may function as an early warning signal or entry point into the depression–precariousness cycle, making it a promising target for preventive intervention. 
In contrast, social precariousness more often appeared as an effect rather than a cause—receiving arrows from symptoms like depressed mood and guilt, but rarely sending them. This asymmetry implies that social precariousness may often reflect downstream consequences of worsening mental health rather than being its root cause [@lund2010poverty; @compton2015social]. While such relationships are complex and likely reciprocal over time, the directional patterns observed provide a useful foundation for hypothesis generation and targeted intervention.

>>>>

```{r, include=FALSE, cache=TRUE}
# Load the dataset 
dat <- read_sav("../data/helius_dat/HELIUS_LEONIE.sav")

@@

### Depression as sum score
@fig-sum provides a high-level summary of how precariousness factors collectively influence and are influenced by overall depression severity, focusing on the causal relationships between precariousness factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*), as identified by the FCI algorithm.
<!-- As described in [@sec-analysis], dashed edges indicate ties, and the edge endpoint at which the tie occurs is marked by a bold horizontal bar. -->

<!-- The causal graph in panel (a) highlights key pathways in the relationships between precariousness factors and depression (*PHQsum*).  -->
In panel (a) we see that employment precariousness (*P.emp*) and social precariousness (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role. While *P.emp* and *P.soc* are generally not recognized as causes of other precariousness factors, *S.fin* emerges as a potential cause, as indicated by its circle edge endpoint. This pattern is further supported by the edge-type proportion matrix shown in panel (b), which aggregates edge inference outcomes across bootstrapped datasets.

<!-- The matrix plot can be interpreted such that the symbol in matrix[$i$, $j$] represents the relationship $i -~[symbol] \  j$. For example, if matrix[$i$, $j$] is > and matrix[j, i] is $\circ$, then the inferred relationship is $i \circarrow j$.  In the table, different edge types are represented by distinct colors: light gray for the absence of an edge, blue for circles, green for arrowheads, and coral for arrowtails. The colors are blended based on the proportion of each edge type, with higher proportions increasing the opacity of the corresponding color, making dominant symbols more visually prominent.  -->

@@

The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI analysis, with one key distinction: CCI shows greater ambiguity in the directional relationship between *S.fin* and *PHQsum*: CCI assigns an arrowhead from *S.fin* to *PHQsum* in roughly 50% of bootstraps, while the remaining cases display an undirected (circle) endpoint.
Overall, CCI introduces greater uncertainty in edge orientations relative to FCI, yielding more undirected endpoints and more frequent bidirectional (confounded) edges, which in turn contribute to a slightly higher overall count of arrowheads when directions are inferred. 
<!-- Despite this, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm. For example,  -->
Both algorithms identify directional pathways from the stressors (*S.rel*, *S.fin*) to *P.soc* and suggest that *P.soc* is not a cause of *PHQsum*, but more likely affected by it. *P.emp* is not inferred to cause any of its neighboring variables (*P.hou*, *PHQsum*, *S.fin*) and instead appears to be affected by them. *P.hou*, in turn, is connected to *P.emp* and *S.fin* with circle endpoints, indicating that it may be a potential source of influence rather than an outcome. 
<!-- These consistent pathways underscore the robustness of the core relational architecture across algorithms amid variation in directional confidence. : " This is discussion point : Vitor-->

@@

Causal relationships between five precariousness factors and depression sum score (PHQsum), as estimated using the FCI algorithm across bootstrapped samples.
(a) Aggregated causal graph showing the most frequent edge types across bootstraps. Dashed edges indicate ties, with a bold horizontal bar marking where the tie occurs.
(b) Proportion matrix summarizing edge types across all bootstraps. Each matrix cell $[i, j]$ represents how often an edge from variable $i$ to variable $j$ exhibited a particular edge mark—arrowhead, tail, circle, or no edge—adjacent to $j$. The edge mark in matrix$[i,j]$ reflects the orientation of the edge ending at $j$. For example, if matrix$[i, j]$ shows a green arrowhead ($>$) and matrix$[j, i]$ shows a blue circle ($\circ$), then the inferred relationship is $i \circarrow j$. Edge types are encoded by distinct colors: green for arrowheads, coral for arrowtails, blue for circles, and light gray for absence of an edge. Colors are blended and shaded by frequency, with greater opacity indicating higher confidence in the inferred edge type.

@@

Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms. Unlike the sum score graph, which aggregates all symptoms into a single measure, the symptom-level graph in @fig-sym reveals the heterogeneity in how specific symptoms (*con* (concentration), *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood)) are influenced by, and in turn influence, different forms of precariousness.

@fig-sym-1 reveals a strong interdependence among symptoms and suggests much presence of latent confounders, as indicated by numerous bidirectional edges. 
<!-- As before, dashed edges denote ties, with specific ties marked by bold horizontal bars.  -->
In @fig-sym-2, the proportion matrix indicates that arrowheads are the most frequently inferred edge type. However, circle endpoints are particularly common for *anh*, *slp*, *ene*, and *sui*, reflecting uncertainty in the causal direction among these symptoms. Notably, the only arrowtail connection appears between *dep* and *anh*, suggesting that anhedonia is a potential cause of depressed mood.

@@

<!-- This discrepancy likely arises from a combination of factors.  -->
<!-- As the analysis moves from aggregated to disaggregated variables, the number of CI tests increases substantially, leading to a loss of statistical power. With more variables being conditioned on, the effective sample size per test diminishes, making it harder to detect subtle dependencies. Over-conditioning—controlling for too many variables—can also obscure real relationships, resulting in false negatives. -->
<!-- Beyond these statistical power issues, the underlying assumptions of the two CI tests contribute to their divergence. RCoT is a nonparametric test that captures both linear and nonlinear dependencies without assuming a specific distributional form. However, it is more conservative and requires larger sample sizes, especially in high-dimensional settings. In contrast, the Gaussian CI test is based on partial correlations and assumes linear Gaussian relationships, making it more permissive and more likely to detect weak or spurious associations. -->
<!-- As a result, it is common to see edges where Gaussian CI suggests a directional relationship, while RCoT indicates no edge. In the proportion matrix, this appears as a 50–50 tie between absence and arrowhead symbols.  -->
<!-- This systematic divergence illustrates the broader challenge of causal discovery in high-dimensional data: different statistical assumptions can lead to different conclusions about whether—and in which direction—variables are causally related. -->

@@

Other precariousness factors exhibit more uncertainty but still notable relationships. *S.rel* is connected to *slp*, *glt*, and *app*, though directionality remains unclear in many cases. *P.emp* also connects to symptoms like *slp*, *mot*, and *con*, but, like *S.rel*, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
In line with the aggregated graph in @fig-sum, *P.hou* does not appear to have any direct relationship with depression symptoms but consistently shows associations with *P.emp* and *S.fin*. The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with *S.fin* exhibiting the strongest tendency to influence other precariousness factors.

Among depressive symptoms, *dep*, *glt*, and *slp* appear to be the most connected to precariousness factors, while *P.soc* has the most connections with symptoms, predominantly as a recipient rather than a driver of influence. Within the symptom network, *slp*, *sui*, and *anh* emerge as causally influential symptoms, as they exhibit more outgoing arrows compared to other symptoms. On the other hand, *con*, *mot*, *dep*, and *glt*, despite having high connectivity, predominantly receive incoming arrows, indicating they are more likely effects rather than causes.
Considering both symptom-precariousness connections and symptom-level dynamics, *slp* emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network.
<!-- This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precariousness.  -->
*P.soc* appears to function as a bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precariousness, reinforcing a self-sustaining dynamic between depression and precarious conditions.

@@

Unlike the disaggregated graph in @fig-sym, the aggregated symptom-precariousness graph (@fig-presum-1) shows greater certainty in causal directions, with ties occurring only in edges involving the overall precariousness factor.
This is even more evident in the proportion matrix (@fig-presum-2), where symptom-to-symptom interactions almost fully converge to a single edge type. Additionally, all symptom interactions are connected either through bidirectional edges or a combination of an arrowhead and a circle, suggesting a strong presence of latent confounders.

Examining the symptom-precariousness connections, we observe a stronger overall trend of depressive symptoms influencing precariousness, rather than the other way around. Specifically, *anh*, *dep*, and *slp* frequently emerge as sources of influence on precariousness, whereas *app*, *glt*, and *mot* exhibit more uncertainty in directionality, often resulting in circle endpoints. This pattern suggests that when precariousness factors are aggregated, the dominant causal flow is from depressive symptoms to precariousness, rather than precariousness driving depression. Particularly, *dep* emerges as the strongest predictor of precariousness, exhibiting the highest proportion of arrowtails, consistent with findings from the disaggregated analysis.

@@

Across multiple levels of analysis, our results point to a consistent and directional link between precariousness and depression. Financial stress (*S.fin*) emerges as a particularly robust and modifiable upstream driver: in both aggregated and disaggregated models, it shows strong and stable causal influence on depressive symptoms. By contrast, other forms of precariousness—particularly social precarity—more often appear as outcomes rather than antecedents of depression, frequently receiving rather than sending causal arrows. Among symptoms, *dep* and *slp* stand out for their centrality and connections to precariousness factors, with *dep* also implicated in potential feedback loops with social precarity (*P.soc*). These findings motivate the use of a simplified three-node model in the simulation—capturing depression, financial stress, and social precarity—as it preserves the key dynamic motifs observed in the causal graphs while enabling tractable exploration of system-level responses to stress-reducing interventions.

<!-- Overall, the aggregated precariousness graph supports the key findings from both the disaggregated analysis and the depression sum score analysis, providing greater certainty in several important patterns. The results highlight the significant role of latent confounders in symptom-to-symptom interactions. While certain stressor-related precariousness factors, particularly *S.fin* and *S.rel*, appear to causally influence depressive symptoms, other precariousness factors are more likely driven by depression rather than acting as primary causes, suggesting that precariousness may function more as a consequence of depression than its source. Some symptoms, particularly *glt*, *slp*, and *dep*, appear to influence precarious conditions while also occupying central roles within the symptom network. Among them, *slp* and *glt* emerge as influential symptoms within the depression network, whereas *con* and *dep*, despite their high connectivity, behave more as outcomes rather than primary drivers. Together, these findings suggest that depression—especially symptoms like sleep disturbances and guilt—may contribute to sustaining precarious conditions, possibly forming a reciprocal cycle between depression and precariousness. -->

```{r}
#| layout-nrow: 2
#| label: fig-presum
#| fig-cap: "Resulting graph of precariousness sum score and individual depression symptoms using FCI and proportion of edge endpoint types. As before, dashed edges denote ties, with specific ties marked by bold horizontal bars."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 60%

@@

To assess how internal structure influences the system’s responsiveness to external support, we simulate a set of linear models that are consistent with the data but vary in the assumed strength of influence from precariousness to depression ($\alpha_{DP}$).
For each model, we generate a virtual population of 1,000 individuals and simulate an external intervention that reduces each individual’s financial stress ($S$) proportionally toward the minimum observed value in the dataset. While individuals begin with different baseline stress levels, the intervention applies a shared proportional reduction across the population. This intervention defines the external input, while variation in $\alpha_{DP}$ reflects different assumptions about internal feedback from precariousness to depression.
We examine system behavior in two complementary ways: (1) by analyzing average levels of depression and precariousness at selected time points following intervention onset, and (2) by simulating full temporal trajectories under sustained intervention followed by recovery, using two representative cases at the extremes of the $\alpha_{DP}$ range.

@fig-intervention summarizes how variation in $\alpha_{DP}$ shapes system-level responses to external support.
<!-- Panel A shows the mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, with the x-axis indicating the strength of external support, defined as the proportional reduction of $S$ toward its minimum observed value. -->
<!-- Each line represents a different scenario with a different $\alpha_{DP}$, color-coded from low (purple) to high (yellow). -->
In Panel A, during the early phase of intervention, depression and precariousness exhibit contrasting patterns. Depression declines more rapidly in scenarios with lower $\alpha_{DP}$, where its dynamics are driven more directly by the external stressor $S$. 
As $\alpha_{DP}$ increases, $D$ becomes more dependent on precariousness ($P$), which itself adjusts more slowly—delaying the depressive response.
Precariousness, by contrast, shows slightly faster early reductions in scenarios with higher $\alpha_{DP}$. This reflects a shift in influence: as $\alpha_{DP}$ increases, $P$ becomes more directly responsive to $S$ (via increasing $\alpha_{PS}$) and less influenced by $D$, allowing it to adjust more promptly to changes in the stressor.

Despite these short-term differences, all scenarios ultimately converge to similar steady-state levels under sustained intervention. AS the systm approaches convergence, some variation remains, particularly in $P$, where gradients between models are still visible. 
<!-- By step 500, trajectories have nearly fully aligned across all configurations, consistent with the shared equilibrium structure of the linear system.  -->
This convergence highlights that the key distinctions lie not in whether intervention is effective, but in how quickly and through which pathways change unfolds.

Panel B shows full time trajectories for two representative cases: one with minimal internal feedback ($\alpha_{DP} = 0$) and one at the upper admissible limit ($\alpha_{DP} = 0.65$). In both, there is a temporary intervention period. When $\alpha_{DP} = 0$, depression responds rapidly to the reduction in stress and rebounds shortly after support is withdrawn. In contrast, when $\alpha_{DP}$ is high, $D$ responds more gradually and recovers more slowly—reflecting its increased dependence on the slower-changing $P$.
Precariousness declines in both scenarios during intervention and returns to baseline afterward, but with a slightly sharper drop in the higher $\alpha_{DP}$ condition. 
This occurs because $P$ becomes more directly responsive to $S$, as consistency with the data requires $\alpha_{PS}$ to increase with $\alpha_{DP}$ (see @fig-tradeoff). These parameter shifts reflect trade-offs needed to preserve the empirical covariances, with further details provided in Appendix @sec-analcalibration.

@@

```{r}
#| label: fig-intervention
#| fig-cap: >
#|   System response to external intervention across variations in internal structure.
#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across varying values of $\alpha_{DP}$ and intervention strengths. The x-axis represents the strength of external support, operationalized as the proportional reduction of financial stress ($S$) toward its minimum observed value. Each line corresponds to a different model defined by a specific $\alpha_{DP}$ value, color-coded from low (purple) to high (yellow). Models with lower $\alpha_{DP}$ exhibit faster early reductions in $D$, while higher $\alpha_{DP}$ leads to more immediate declines in $P$.
#|   
#|   **Panel B**: Full time trajectories for two models—one with no coupling ($\alpha_{DP} = 0$) and one with strong coupling ($\alpha_{DP} = 0.65$)—illustrating system response and recovery following intervention onset (step 300) and withdrawal (step 1500).
#|   Depression responds more gradually and persistently in the high $\alpha_{DP}$ model, whereas precariousness adjusts more quickly.*

@@

This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precariousness interact and how their internal structure modulates responsiveness to intervention.
Consistent with prior work linking financial strain to psychological distress [@lund2010poverty; @ezzy1993unemployment], our causal analysis identified financial stress as a key upstream driver of depression—emerging across both aggregated and disaggregated models. This provides empirical support for longstanding hypotheses that economic stressors can act as modifiable entry points into broader cycles of psychosocial risk.

At the symptom level, our analysis aligns with established theories of psychopathology but offers a more fine-grained map of potential causal relationships, generating concrete, testable hypotheses. It also points to possible limitations of prevailing models, as some observed symptom–symptom links may be shaped by shared confounders rather than direct causal influence. This suggests that, in some contexts, broader mechanisms or clusters of symptoms may provide more effective units of analysis and intervention than individual symptoms alone. 
Within this framework, sleep disturbance emerged as a potentially initiating symptom, with directed connections to both symptom-level and precariousness variables—mirroring longitudinal findings that identify sleep disruption as an early warning signal for depressive episodes and social withdrawal [@zawadzki2013rumination; @alvaro2013systematic].
Depressed mood, another core symptom, occupied a central position in the network. Not only was it influenced by upstream symptoms like anhedonia and guilt, but it also showed directional influence on social precariousness—suggesting that internal psychological states may actively shape external vulnerability. The CCI algorithm further identified a potential feedback loop between depressed mood and overall precariousness, resonating with social drift theories in depression, which posit that declining mental health can lead to social exclusion, job instability, and housing insecurity, reinforcing depressive cycles [@compton2015social; @boschloo2015network].

In contrast, social precariousness itself appeared more often as an effect than a cause, receiving arrows from symptoms like depressed mood and guilt but rarely sending them. This asymmetry supports prior observations that indicators of social vulnerability often reflect downstream consequences of poor mental health, though the relationships are likely bidirectional over time [@lund2010poverty]. 
These patterns, while derived from Amsterdam-specific data, align with mental health research in other high-income city settings, where structural inequalities such as poverty, unemployment, and housing insecurity are persistent drivers of poor psychological outcomes [@fryers2003social; @weich1998poverty].
Although the specific dynamics may vary by context, the parallels with broader literature suggest that similar mechanisms could operate in comparable settings. Nonetheless, replication in other populations will be important for assessing robustness and external validity.

Taken together, these findings yield actionable insights for mental health intervention. Symptoms like sleep disturbance, which demonstrate early and broad influence across both symptom and precariousness networks, may serve as effective targets for prevention—potentially halting the progression into more entrenched depressive states or worsening precariousness. Similarly, intervening on central nodes like depressed mood may offer strategic leverage for breaking mutually reinforcing cycles of mental decline and social isolation.
Importantly, interventions targeting such nodes may have ripple effects across domains. For instance, evidence suggests that treating core depressive symptoms can improve social functioning, employment engagement, and housing stability—particularly when combined with social or financial supports [@mcdaid2019economic; @lund2018social]. These dual benefits underscore the value of integrated, cross-domain strategies that simultaneously address internal symptoms and external stressors.
